---
title: "HLAtools"
author: 
- Ryan Nickens -- rynickens@gmail.com
- Livia Tran -- livia.tran@ucsf.edu
- Leamon Crooms IV -- lcroomsiv@gmail.com
- Derek Pappas, PhD -- djpappas75@gmail.com
- Vinh Luu -- vinhluu864@berkeley.edu
- Josh Bredeweg
- Steven J. Mack, PhD -- steven.mack@ucsf.edu
date: "2024-03-25"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HLAtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HLAtools)
```
-   Package Version: 0.8.1.9000

# Overview
The *HLAtools* package contains a suite of tools and computable resources that facilitate the effective analysis of the named genes in the HLA region.

These include:

- Reference datasets that describe the genes supported by the IPD-IMGT/HLA Database, organize them into useful categories, and identify their structural element boundaries.
- The ability to build local amino acid, codon, coding nucleotide, and genomic nucleotide sequence alignments for HLA genes of interest.
- Advanced tools of the dissection and comparison of allele sequences, and the construction of customized multi-gene sequence alignments.
- Functions that translate Genotype List (GL) String Code messages across IPD-INGT/HLA database release versions, and  inter-convert between GL String and UNIFORMAT formats.
- Data-analysis tools that extend the utility of Bridging Immunogenomic Data-Analysis Workflow Gaps (BIGDAWG) formatted datasets by allowing stratification and calculating Relative Risk. 
  

